Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
add a couple of todos

\paragraph{Making Better Use of the Reward Signal}
Several existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.

\paragraph{Cumulative Regret}
Although we have focused on simple regret in our analysis, a natural but open question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with graph feedback to get optimal cumulative regret guarantees.
Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
\todom{Check this!}
Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.

@@

\paragraph{Causal Models with Non-Observable Variables}
If we assume knowledge of the conditional \textit{interventional} distributions $P_a$ our analysis applies unchanged to the case of causal models with non-observable variables. For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$. Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from even an infinite amount of observational data. \todof{tie in to brief discussion of approach if had to estimate $P_a$ terms within algorithm.

>>>>

\paragraph{Making Better Use of the Reward Signal}
Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
For the parallel bandit the required modifications should be quite straightforward. The idea would be to adapt the algorithm to essentially use successive elimination in
the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability. In the general case a similar modification is also possible
by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping. Note
that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal. For this reason we prefer to emphasize 
the main point that causal structure should be exploited when available. Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which
in some cases may be preferred to a sequential design for logistical reasons. This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.

\paragraph{Cumulative Regret}
Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information 
to get near-optimal cumulative regret guarantees. They consider a finite-armed bandit model with side information where in reach round
the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
In this way the learner can gain information about actions it does not take with varying levels of accuracy. The reduction follows by substituting
the importance weighted estimators in place of the Gaussian reward. In the case that $\vec{q}$ is known this would lead to a known variance
and the only (insignificant) difference is the Bernoulli noise model. In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
at least asymptotically.

%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
%\todom{Check this!}
%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.

@@

\paragraph{Causal Models with Non-Observable Variables}
If we assume knowledge of the conditional \textit{interventional} distributions $P_a$ our analysis applies unchanged to the case of causal models with non-observable variables. For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$. Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from even an infinite amount of observational data. \todof{tie in to brief discussion of approach if had to estimate $P_a$ terms within algorithm.}

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
adds truncated product formula
experiments thanks to AWS

\subsection{Parallel bandits problem}

For the parallel bandit problem, we use a model in which, unknown to the algorithms, $Y$ depends only on a single variable.
\eq{
Y_t \sim \begin{cases}
\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
\end{cases}
}

This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\left(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon\right)$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions. We set $\boldsymbol{q} = 0$ for $q_i \leq m$ and $\frac{1}{2}$ otherwise. Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.

@@

In figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases. The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure. For the causal algorithms it increases approximately with $\sqrt{m}$. As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s1000_20160205_0910.pdf}
\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
\label{fig:simple_vs_m}
\end{figure}

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\epsilon = \sqrt{\frac{N}{8T}}$ , is just too small to be learned. This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m << N$. In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

@@

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
\label{fig:simple_vs_T_vary_epsilon}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
\label{fig:simple_vs_T}
\end{figure}

>>>>

We compare Algorithms \ref{alg:simple} and \ref{alg:general} with Successive Elimination on the parallel bandit problem
under a variety of conditions, including where the importance weighted estimator used by Algorithm \ref{alg:general} is not truncated,
which is justified in this setting by Remark \ref{rem:truncate}. Throughout we use a model in which $Y$ depends only on a single variable (this is unknown to the algorithms). 
\eq{
Y_t \sim \begin{cases}
\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
\end{cases}
}
This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\left(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon\right)$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions. We set $\boldsymbol{q} = 0$ for $q_i \leq m$ and $\frac{1}{2}$ otherwise. Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.

@@

In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases. The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure. For the causal algorithms it increases approximately with $\sqrt{m}$. As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
\label{fig:simple_vs_m}
\end{figure}

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\epsilon = \sqrt{\frac{N}{8T}}$ , is just too small to be learned. This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m << N$. In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

@@

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1316}
\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
\label{fig:simple_vs_T_vary_epsilon}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1303}
\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
\label{fig:simple_vs_T}
\end{figure}

### 2016-02-05 causal-bandit/icml-2015/general.tex
add a couple of todos

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
  	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
\caption{A causal chain graph}.
\label{fig:causalchain}
\end{figure}

Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
a truncated importance weighted estimator.

Let $a \in \calA$ be an intervention and define random variable
\eq{
Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
}
where $B_a \geq 0$ is some constant to be chosen subsequently and
\todom{Symbol clash with regret}
\eq{
R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\,.
}
Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}
We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE {\bf return} $\hat a^* = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

@@

@@

The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
Recall in that setting that 
\eq{
\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
}
For $a = do()$ choose $\eta_a = 1/2$. 
\todom{Check please!}
Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.

>>>>

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
  	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
\caption{A causal chain graph}.
\label{fig:causalchain}
\end{figure} 
\todof{connect truncated product formula to here by noting that the truncated formula allows us to estimate $P_a$ from observational quantities - which could plausible be available prior to running the experiment, allowing us to assume the $P_a$'s are known.}

Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
a truncated importance weighted estimator. The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail
that is so detrimental to its concentration guarantees.

Let $a \in \calA$ be an intervention and define random variable
\eq{
Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
}
where $B_a \geq 0$ is some constant to be chosen subsequently and
\eq{
R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\,.
}
Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}
We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately 
coincides with the version for parallel bandits, justifying the name overloading.

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

@@

\begin{remark}\label{rem:truncate}
The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
often possible to choose $B_a$ significantly
larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
to learn from observational data.
\end{remark}

@@

The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
Recall in that setting that 
\eq{
\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
}
For $a = do()$ choose $\eta_a = 1/2$. 
Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
